Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-04-18 Paper/Paper.tex

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,hyperref,array,xcolor,multicol,verbatim,mathpazo}
\usepackage[normalem]{ulem}
\usepackage[pdftex]{graphicx}
\usepackage{fullpage}
\usepackage{import}
\usepackage{adjustbox}
\usepackage{booktabs}
\usepackage{bbm}
\usepackage[font=normalsize,labelfont=bf]{caption}
%\captionsetup{justification=raggedright,singlelinecheck=false}
\usepackage[english,dutch]{babel}

@@

Now, I examine how these result differ when considering physicians in different age brackets: pre-retirement and typical retirement age. These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the result in the first year after EHR exposure. A physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean. In comparison, physicians less than 60 are far less likely, if at all, to respond to EHR exposure in the first year after EHR exposure. Physicians in the young category have a larger effect on the likelihood of retiring in the second year after exposure relative to the first year. While I do not have data to test this, if young physicians are more likely to switch careers after EHR exposure then it may take one additional year to prepare for that decision relative to retirement.

@@

\subsection{Patient Count}\label{sec:patientcount}

An important, and heavily debated, aspect of the implementation of EHRs is whether or not they increase productivity in users. I measure productivity as the number of Medicare patients seen annually. EHRs could increase this number if they improve efficiency through less paperwork or ease of information access, or EHRs could decrease the number of patients seen if physicians are spending more time interfacing with the system and are not able to see as many patients.

The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}. For physicians of any age, patient count increases by 11.5 patients in the year of EHR exposure, 24 patients in the year after exposure, 15.3 patients in the second year after exposure, and then no effect after the second year. These results translate to a .5-1\% increase in the number of patients seen relative to the mean. As with other outcomes, this result is totally due to changes by the younger group of physicians. Senior physicians, who are on average likely more averse to technology, are not gaining in efficiency due to EHR use whereas younger physicians, who may pick up the technology quicker, experience efficiency gains immediately.

@@

\subsection{Billing Activity}

@@

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Retirement}
    \includegraphics[scale=.57]{Objects/retire_plot.pdf}
    \label{fig:retirefirst}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
    \includegraphics[scale=.57]{Objects/officeind_plot.pdf}
    \label{fig:officefirst}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
    \includegraphics[scale=.57]{Objects/officefrac_plot.pdf}
    \label{fig:officesecond}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
    \includegraphics[scale=.57]{Objects/zip_plot.pdf}
    \label{fig:zip}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Patient Count}
    \includegraphics[scale=.57]{Objects/patient_plot.pdf}
    \label{fig:patient}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Claim Count}
    \includegraphics[scale=.57]{Objects/claim_plot.pdf}
    \label{fig:claim}
\end{figure}

>>>>

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,hyperref,array,xcolor,multicol,verbatim,mathpazo}
\usepackage[normalem]{ulem}
\usepackage[pdftex]{graphicx}
\usepackage{fullpage}
\usepackage{import}
\usepackage{adjustbox}
\usepackage{booktabs}
\usepackage{bbm}
\usepackage[font=normalsize,labelfont=bf]{caption}
%\captionsetup{justification=raggedright,singlelinecheck=false}

@@

Next, I examine how these result differ when considering physicians in different age brackets: pre-retirement and typical retirement age. These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the result in the first year after EHR exposure. A physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean. In comparison, physicians less than 60 are far less likely, if at all, to respond to EHR exposure in the first year after EHR exposure. Physicians in the young category have a larger effect on the likelihood of retiring in the second year after exposure relative to the first year. While I do not have data to test this, if young physicians are more likely to switch careers after EHR exposure then it may take one additional year to prepare for that decision relative to retirement.

@@

\subsection{Patient Count and Billable Activity}\label{sec:patientcount}

An important, and heavily debated, aspect of the implementation of EHRs is whether or not they increase productivity in users. A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden. Such a decrease implies a change in patient care: either more patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments. In this section, I estimate the effect of EHR implementation on a physician's number of patients seen.

The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}. For physicians of any age, patient count increases by 11.5 patients in the year of EHR exposure, 24 patients in the year after exposure, 15.3 patients in the second year after exposure, and then no effect after the second year. These results translate to a .5-1\% increase in the number of patients seen relative to the mean. As with other outcomes, this result is totally due to changes by the younger group of physicians. While there is a statistically larger effect for younger physicians relative to senior physicians (who likely take longer to adapt to technology changes), the observed effect is economically trivial.

Since this is a demand-driven variable as well, I investigate whether any external factors may shift the demand for health care over the sample period.

@@

@@

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Retirement}
    \includegraphics[scale=.4]{Objects/retire_plot.pdf}
    \label{fig:retirefirst}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
    \includegraphics[scale=.4]{Objects/officeind_plot.pdf}
    \label{fig:officefirst}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
    \includegraphics[scale=.4]{Objects/officefrac_plot.pdf}
    \label{fig:officesecond}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
    \includegraphics[scale=.4]{Objects/zip_plot.pdf}
    \label{fig:zip}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Patient Count}
    \includegraphics[scale=.4]{Objects/patient_plot.pdf}
    \label{fig:patient}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Claim Count}
    \includegraphics[scale=.4]{Objects/claim_plot.pdf}
    \label{fig:claim}
\end{figure}
